Interactive Checks for Coordination Avoidance

### 2019-08-18 doc/vldb_slides/sections/bank_example.tex
Tweaks, plus gutted poem.

  \note{%
    So, what is invariant confluence? Well, let's begin with an example.

@@

    Users can issue transactions to the database, and every transacation is
    either a deposit or a withdrawal. And, I have one critical invariant that I
    want to maintain over my database. And that invariant is that my bank
    account balance should be non-negative at all times. I'm not allowed to go
    in the hole.
  }
\end{frame}

@@

  \note{%
    Now, instead of implementing our bank with strong consistency, we could
    instead implement it with weak consistency and avoid coordination
    altogether.

@@

  \note{%
    Now, these are just two examples, but if you think about it long enough,
    they are representative of a trend. We can implement deposit transactions
    without any coordination and we're guaranteed to never violate the
    invariant. Withdrawal transactions, on the other hand, do require
    coordination to avoid violating the invariant.

    At a very high level, this is exactly the notion that invariant confluence
    captures. We'll see that deposit transactions are invariant confluent and
    thus do not require coordination while withdrawal transactions are not
    invariant confluent and thus do require coordination.
  }
\end{frame}

>>>>

  \note{%
    We'll begin with an example.

@@

    Users can issue transactions to the database, and every transacation is
    either a deposit or a withdrawal. And, I have one critical invariant that I
    want to maintain over my database. And that invariant is that my bank
    account balance should never be negative. My balance must be non-negative at all times.
  }
\end{frame}

@@

  \note{%
    Alternatively, instead of implementing our bank with strong consistency, we
    could implement it with weak consistency and avoid coordination altogether.

@@

  \note{%
    We can learn a lot from these examples.

    First, if we implement our bank with strong consistency, we never violate
    our invariant.

    Second, we can implement deposit transactions without any coordination and
    we're still guaranteed to never violate the invariant.

    Third, withdrawal transactions, unlike deposit transactions, do require
    coordination to avoid violating the invariant.

    At a very high level, invariant confluence tries to capture exactly which
    transactions require coordination to maintain invariants and which do not.
    In this example, deposit transactions are invariant confluent and thus do
    not require coordination while withdrawal transactions are not invariant
    confluent and thus do require coordination.
  }
\end{frame}

### 2019-08-18 doc/vldb_slides/sections/four_wise_men.tex
Tweaks, plus gutted poem.

\begin{frame}
  \note{%
    Hi everyone! Thanks so much for attending. I'm Michael, and today I'll be
    talking about ``Interactive Checks for Coordination Avoidance''. This is
    joint work with my advisor at UC Berkeley, Joe Hellerstein.

@@

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{You should use weak consistency. It's wicked fast!}
      };
      \path[fill=gray!30] (vogels.north) -- ++(0, 1) -- ++(0.5, 0) -- (vogels.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Someone like Werner Vogels would probably tell you to use weak consistency.
    You can implement weak consistency super duper fast, and for many
    applications, weak consistency is good enough.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \node[quote, text width=0.8\textwidth, align=left] at (1, 3) {
        \Large\textit{What about the application invariants?!}
      };
      \path[fill=gray!30] (hoare.north) -- ++(0, 1.5) -- ++(1, 0) -- (hoare.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    But then Tony Hoare laments, ``no no no''. Hoare encourages us not to
    settle for weak consistency because a weakly consistent system is hard to
    reason about. It robs us of a fundamental way we use to reason about
    programs: application invariants. Even if every transaction in a weakly
    consistent system preserves application invariants, the whole system may
    not.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \node[quote, text width=0.8\textwidth, align=left] at (2, 3) {
        \Large\textit{Yeah, go with strong consistency! I have an algorithm for that :)}
      };
      \path[fill=gray!30] (lamport.north) -- ++(0, 1.5) -- ++(1, 0) -- (lamport.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Leslie Lamport chimes in agreeingly. He encourages us to choose strong
    consistency. Strongly consistent systems are much easier to reason about.
    After all a strongly consistent system to end users looks exactly like a
    system that's not even distributed.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \brewer
      \node[quote, text width=0.8\textwidth, align=left] at (3, 3) {
        \Large\textit{Not so fast!}
      };
      \path[fill=gray!30] (brewer.north) -- ++(0, 1.5) -- ++(-1, 0) -- (brewer.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    ``Aha, not so fast'' counters Eric Brewer! Consistency and availability and
    partition tolerance, oh my. Surely, strong consistency comes at the price
    of availability and performance.

    So we're in a bit of a pickle. Weak consistency is too confusing and strong
    consistency is too heavy handed.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \node[pic] (bailis) at (0, 0) {\includegraphics[width=0.25\textwidth]{assets/bailis.jpg}};
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{?`Porque no los dos?}
      };
      \path[fill=gray!30] (bailis.north) -- ++(0, 1.5) -- ++(1, 0) -- (bailis.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Thankfully, Peter Bailis comes to save the day and reminds us that perhaps
    we can have the best of both worlds. In 2014, Peter and some other smart
    folks defined invariant confluence as a condition under which we can
    replicate a piece of data without the need for strongly consistent
    coordination while maintaining the data's application invariants.

    Invariant confluence will be the topic of this talk. In particular, I'll
    define invariant confluence and then discuss how we can check invariant
    confluence automatically.
  }
\end{frame}

>>>>

\begin{frame}
  \note{%
    Hi everyone! Thanks so much for attending. I'm Michael, and today I'll be
    talking about ``Interactive Checks for Coordination Avoidance''. This is
    joint work with my advisor at UC Berkeley, Joe Hellerstein. \\[12pt]

    % I'd like to begin my talk with a short fairy tale. The fairy tale involves
    % an application developer who has some data they need to replicate.
    % Unfortunately, the application developer cannot decide whether to replicate
    % with weak consistency or strong consistency. The application developer
    % stays up all night trying to decide but eventually falls asleep at their
    % desk. In their sleep, they are visited by four wise men. And so, the fairy
    % tale begins:

@@

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{
        Go with weak consistency. It's super duper fast and good enough for
        many applications!
        % Weak. Weak. It's weak that you're seeking.
        % \pause
        % Please learn that if you yearn your performance to peak, then turn no
        % further than weak.
      }
      };
      \path[fill=gray!30] (vogels.north) -- ++(0, 1) -- ++(0.5, 0) -- (vogels.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Someone like Werner Vogels would probably tell you to use weak consistency.
    You can implement weak consistency super duper fast, and for many
    applications, weak consistency is good enough. \\[12pt]

    % ``Weak. Weak. It's weak that you're seeking'', squeaked Werner Vogels.
    % His voice trembled, creaking, as he kept speaking.
    % ``Please learn that if you yearn your performance to peak,
    % then turn no further than weak.'' critiqued Werner.
    %
    % Werner Vogels has a good point. You can implement weak consistency super
    % duper fast, and for many applications, weak consistency is good enough.
    % But, let's continue.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \node[quote, text width=0.8\textwidth, align=left] at (1, 3) {
        \Large\textit{
          Weakly consistent systems are so hard to reason about!
          % No. No. No more.
          % \pause
          % Weak is not easy. Weak is just treason. Please, these anomalies make
          % it so hard to reason.
        }
      };
      \path[fill=gray!30] (hoare.north) -- ++(0, 1) -- ++(0.5, 0) -- (hoare.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Tony Hoare would encourage us not to settle for weak consistency because a
    weakly consistent system is hard to reason about.

    % ``No, no, no more'' groaned Tony Hoare.
    % He moaned, he lamented, his soul was tormented
    % ``Weak is not easy. Weak is just treason.
    % Please, these anomalies, make it so hard to reason.''
    %
    % You see, Tony Hoare encourages us not to settle for weak consistency
    % because a weakly consistent system is very hard to reason about.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \node[quote, text width=0.8\textwidth, align=left] at (2, 3) {
        \Large\textit{
          Go with a strongly consistent system! I think there are some good
          algorithms for that kind of stuff.
          % You see, I agree.
          % \pause
          % Weak is just wrong, not right for long. To defend your end users, you
          % must go with strong.
        }
      };
      \path[fill=gray!30] (lamport.north) -- ++(0, 0.5) -- ++(0.25, 0) -- (lamport.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Leslie Lamport would agree. He would encourage us to choose strong
    consistency. Strongly consistent systems are much easier to reason about.
    After all a strongly consistent system to end users looks exactly like a
    system that's not even distributed.

    % ``You see, I agree'', says he, the ever mesmerizing Leslie.
    % ``Weak is just wrong, not right for long.
    % To defend your end users, you must go with strong.''
    %
    % Leslie Lamport encourages us to choose strong consistency becausesStrongly
    % consistent systems are much easier to reason about. After all a strongly
    % consistent system to end users looks exactly like a system that's not even
    % distributed.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \vogels
      \hoare
      \lamport
      \brewer
      \node[quote, text width=0.8\textwidth, align=left] at (3, 3) {
        \Large\textit{
          Not so fast!
          % Alas, not so fast!
          % \pause
          % Consistency weak but not weary, strong but not nearly as fast as you
          % think when you hear my CAP theory.
        }
      };
      \path[fill=gray!30] (brewer.north) -- ++(0, 1) -- ++(-1, 0) -- (brewer.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    Eric Brewer would counter that strong consistency isn't perfect. Strong
    consistency comes at the price of availability and performance. So we're in
    a bit of a pickle. Weak consistency is too confusing and strong consistency
    is too heavy handed.

    %
    % ``Alas, not so fast'', Brewer the dapper chap clapped back.
    % ``Consistency weak but not weary, strong but not nearly
    % as fast as you think when you hear my CAP theory.''
    %
    % Eric Brewer is right. Surely, strong consistency comes at the price of
    % availability and performance. So our application developer is in a bit of a
    % pickle. Weak consistency is too confusing and strong consistency is too
    % heavy handed. Fortunately, a hero emerges to save the day.
  }
\end{frame}

\begin{frame}
  \begin{center}
    \begin{tikzpicture}[xscale=1.3]
      \node[pic] (bailis) at (0, 0) {\includegraphics[width=0.25\textwidth]{assets/bailis.jpg}};
      \node[quote, text width=0.8\textwidth, align=left] at (0, 3) {
        \Large\textit{Why not have both?}
      };
      \path[fill=gray!30] (bailis.north) -- ++(0, 1.5) -- ++(1, 0) -- (bailis.north);
    \end{tikzpicture}
  \end{center}

  \note{%
    % Out teetered Peter, with swagger demeanor.
    % ``Why not have both?'' quoth the Peter.
    % The end.

    Fortunately, Peter Bailis can save the and reminding us of when in 2014, he
    and some other smart folks including Alan defined invariant confluence as a
    way to get the benefits of both weak and strong consistency. An invariant
    confluent object can be replicated without coordination while maintaining
    its application invariants.

    Invariant confluence will be the topic of this talk, so let's build some
    intuition on what invariant confluence is.
  }
\end{frame}

### 2019-08-18 doc/vldb_slides/sections/iclosure.tex
Finished first cut at speaker notes.
Tweaks, plus gutted poem.

  \note{%
    Unfortunately, developing an invariant confluence decision procedure
    straight up is not easy. Invariant confluence is fundamentally a property
    about reachable states but reasoning automatically about reachable states
    is hard. I'm not saying it's impossible, but certainly I don't know how to
    do it.

@@

  \note{%
    First, a quick refresher on terminology. We say a set $S$ is closed under a
    binary operator $f$ if for every $x$ and $y$ in $S$, $f(x, y)$ is also in
    $S$. For example, even numbers are closed under addition because the sum of
    any two even numbers is even. But odd numbers are not closed under addition
    because the sum of two odd numbers is even.
  }
\end{frame}

@@

  \note{%
    Invariant closure is a sufficient condition for invariant confluence. Why
    is that? Well, transactions already preserve invariants. So if merging does
    as well, then every reachable state always maintains the invariant.
  }
\end{frame}

@@

  \note{%
    The good thing about invariant closure is that it's much easier to check
    automatically. for example, remember that example we had with pairs of
    integers? We can pose whether or not that object is invariant closed as
    this formula, and we can pass this formula directly to an SMT solver to
    figure out whether it's invariant closed.
  }
\end{frame}

@@

  \note{%
    So to decide whether an object is invariant confluent, we can take the
    object and ask if it's invariant closed. If it is, then it's also invariant
    confluent and we're done. If it's not invariant closed, then what do we
    know?
  }
\end{frame}

@@

  \note{%
    To see if it does, let's revisit the example from before. Recall that in
    this example, our object is invariant confluent.
  }
\end{frame}

@@

  \note{%
    But, note that our object is not invariant closed. The points $(-1, 2)$ and
    $(2, -1)$ both satisfy the invariant, but their merge does not.
  }
\end{frame}

@@

  \note{%
    This example is actually representative of the overall trend. An invariant
    confluent object may not be invariant closed but only because we're merging
    points that are unreachable. If the invariant is a subset of the reachable
    states, then invariant closure and invariant confluence are identical.

    Why? Well, the forward direction here is what we proved before. For the
    backwards direction, we have that the invariant and the set of reachable
    points are identical. The set of reachable states is closed under merge, so
    so is the invariant, so the object is invariant closed.

    I know I'm walking through a lot of theory pretty quickly, but if you
    remember one slide, remember this one. If the invariant is a subset of the
    reachable states, then invariant closure and invariant confluence are
    identical.
  }
\end{frame}

>>>>

  \note{%
    Unfortunately, developing an invariant confluence decision procedure
    straight up is not easy. Invariant confluence is fundamentally a property
    about reachable states but reasoning automatically about reachable states
    is hard.

@@

  \note{%
    First, a quick refresher on what it meas for a set to be closed. We say a
    set $S$ is closed under a binary operator $f$ if for every $x$ and $y$ in
    $S$, $f(x, y)$ is also in $S$. For example, even numbers are closed under
    addition because the sum of any two even numbers is even. But odd numbers
    are not closed under addition because the sum of two odd numbers may not
    be odd.
  }
\end{frame}

@@

  \note{%
    Invariant closure is a sufficient condition for invariant confluence. Why
    is that? Well for an object to be invariant confluent, every reachable
    state must satisfy the invariant. A state is reachable if it can be reached
    by executing transactions merging states. Transactions are already
    guarnateed to preserve invariants, so if merging is also guaranteed to
    preserve the invariant, then every reachable state always satisfies the
    invariant.
  }
\end{frame}

@@

  \note{%
    The good thing about invariant closure is that it's much easier to check
    automatically. Remember that example we had with pairs of
    integers? We can pose whether or not that object is invariant closed as
    this formula, and we can pass this formula directly to an SMT solver to
    figure out whether it's invariant closed. This is not something we can do
    with invariant confluence because it's hard to express reachability as a
    simple formula like this one.
  }
\end{frame}

@@

  \note{%
    To decide whether an object is invariant confluent, we can take the
    object and ask if it's invariant closed. If it is, then it's also invariant
    confluent and we're done. If it's not invariant closed, then what do we
    know?
  }
\end{frame}

@@

  \note{%
    To see if it does, let's revisit the example from before. Recall that in
    this example, our object is invariant confluent. The set of reachable
    states is a subset of the invariant.
  }
\end{frame}

@@

  \note{%
    But, note that our object is not invariant closed. The points $(-1, 2)$ and
    $(2, -1)$ both satisfy the invariant, but if we merge them we get the point
    $(2, 2)$, and the point $(2, 2)$ does not satisfy the invariant.
  }
\end{frame}

@@

  \note{%
    Turns out, this is not a coincidence. An invariant confluent object may not
    be invariant closed but only because we're merging points that are
    unreachable. If the invariant is a subset of the reachable states, then
    invariant closure and invariant confluence are equivalent. \\[12pt]

    Why? Well, the forward direction here is what we proved before. For the
    backwards direction, we have that the invariant is a subset of the set of
    reachable points, and because our object is invariant confluent, the set of
    reachable points is a subset of the invariant. So the invariant and the set
    of reachable points are the same. The set of reachable states is closed
    under merge, so the invariant is also, so the object is invariant closed.
    \\[12pt]

    I know I've walked through a lot of theory pretty quickly, but if you
    remember one slide, remember this one. If the invariant is a subset of the
    reachable states, then invariant closure and invariant confluence are
    equivalent. \\[12pt]
  }
\end{frame}
